---
title: "Part 2 Coursework"
author: '220455224'
date: "`r Sys.Date()`"
output: html_document
---

a)
Load packages
```{r}
library(RSQLite)
library(ggplot2)
```

Establish a connection to the SQLite database
```{r}
conn <- dbConnect(RSQLite::SQLite(), "airline8.db")
```

Create a dataframe for the 10 consecutive years chosen
```{r}
for(i in c(1990:1999)){
  ontime_df <- read.csv(paste0(i, ".csv"), header = TRUE)
  if (i==1990) {
    dbWriteTable(conn, 'ontime', ontime_df,append = TRUE)
  } else {
    dbWriteTable(conn, 'ontime', ontime_df, append = TRUE)
  }
}
```

Find the best time interval of the day to fly for each year.
```{r}
for (year in 1990:1999) {
  # Query to calculate average delay for each time interval of the day for each year
  time_query <- paste("
    SELECT 
        '", year, "' AS Year,
        CASE 
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 0 AND 5 THEN 'Late Night'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 6 AND 11 THEN 'Morning'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 12 AND 17 THEN 'Afternoon'
            ELSE 'Evening'
        END AS time_interval,
        AVG(ArrDelay) AS avg_delay
    FROM ontime
    WHERE ArrTime IS NOT NULL
      AND Year = ", year, "
    GROUP BY time_interval
    ORDER BY avg_delay ASC
  ", sep = "")
  
  # Execute query
  time_result <- dbGetQuery(conn, time_query)
  
  # Print results for each year
  print(time_result)
}
```

Further find the best hour in the morning to fly.
```{r}
for (year in 1990:1999) {
  # Query to calculate average delay for each hourly interval of the morning for each year
  hour_query <- paste("
    SELECT 
        '", year, "' AS Year,
        CASE 
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 6 AND 6 THEN '06:00-07:00'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 7 AND 7 THEN '07:00-08:00'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 8 AND 8 THEN '08:00-09:00'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 9 AND 9 THEN '09:00-10:00'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 10 AND 10 THEN '10:00-11:00'
            WHEN CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 11 AND 11 THEN '11:00-12:00'
            ELSE NULL
        END AS hour_interval,
        AVG(ArrDelay) AS avg_delay
    FROM ontime
    WHERE ArrTime IS NOT NULL
      AND Year = ", year, "
      AND CAST(SUBSTR(DepTime, 1, 2) AS INTEGER) BETWEEN 6 AND 11
    GROUP BY hour_interval
    ORDER BY avg_delay ASC
  ", sep = "")
  
  # Execute query
  hour_result <- dbGetQuery(conn, hour_query)
  
  # Print results for each year
  print(hour_result)
}
```


Find the best day in a week to fly for each year.
```{r}
for (year in 1990:1999) {
  # Query to calculate average delay for each day of the week for each year
  day_query <- paste("
    SELECT 
        '", year, "' AS Year,
        DayOfWeek,
        AVG(ArrDelay) AS avg_delay
    FROM ontime
    WHERE ArrDelay IS NOT NULL
      AND Year = ", year, "
    GROUP BY DayOfWeek
    ORDER BY avg_delay ASC
  ", sep = "")
  
  # Execute query
  day_result <- dbGetQuery(conn, day_query)
  
  # Print results for each year
  print(day_result)
}
```

Find the best month to fly for each year then plot a graph for each year.
```{r}
for (year in 1990:1999) {
  # Query to cslculate average delay for each month of each year
  month_query <- paste("
    SELECT 
        '", year, "' AS Year,
        Month,
        AVG(ArrDelay) AS avg_delay
    FROM ontime
    WHERE ArrDelay IS NOT NULL
     AND Year = ", year, "
    GROUP BY Month
    ORDER BY avg_delay ASC
  ", sep = "")
  
  # Execute query
  month_result <- dbGetQuery(conn, month_query)
  
  # Print results for each year
  print(month_result)
  
  # Convert Month to factor with 12 levels
  month_result$Month <- factor(month_result$Month, levels = 1:12)
  
  # Plot graph for each year
  print(ggplot(month_result, aes(x = Month, y = avg_delay, group = Year, color = Year)) +
    geom_line() +
    geom_point() +
    labs(title = paste("Average Arrival Delay by Month for Year", year),
         x = "Month", y = "Average Arrival Delay") +
    theme_minimal())
}
```
Disconnect from the database
```{r}
dbDisconnect(conn)
```

part b
Import the relevant packages 
```{r}
library(RSQLite)
library(ggplot2)
```

Read and setup table for plane-data.csv
```{r}
planes_df <- read.csv("plane-data.csv", header = TRUE)
dbWriteTable(conn, "planes", planes_df, append = TRUE)
```

Create a dataframe for the 10 consecutive years chosen
```{r}
for(i in c(1990:1999)){
  ontime_df <- read.csv(paste0(i, ".csv"), header = TRUE)
  if (i==1990) {
    dbWriteTable(conn, 'ontime', ontime_df)
  } else {
    dbWriteTable(conn, 'ontime', ontime_df, append = TRUE)
  }
}
```

Create a new dataframe to extract tail numbers of planes with the manufacturing year of 1999 and earlier from planes.csv
```{r}
# Filter out rows where year is not "None"
planes_age <- planes_df[planes_df$year != "None", c("year", "tailnum")]

# Convert year column to numeric
planes_age$year <- as.numeric(planes_age$year)

# Filter planes_age to include only planes made in 1999 and before
planes_age <- subset(planes_age, year <= 1999)

# Rename the "year" column to "manufactured_year"
planes_age <- planes_age %>%
  rename(manufactured_year = year)

# Check the structure of planes_age
str(planes_age)

dbWriteTable(conn,"planes_age_table" , planes_age ,overwrite = TRUE)
```

Merge "ontime" with "planes_age" by the tail number of planes
```{r}
# Execute SQL query to merge ontime with planes_age
merged_df <- dbGetQuery(conn, "
  SELECT 
    ontime.*, 
    planes_age_table.manufactured_year,
    planes_age_table.tailnum AS tailnum_planes
  FROM ontime
  INNER JOIN planes_age_table ON ontime.TailNum = planes_age_table.tailnum
")

# Convert `merged_df` data frame to a table in the SQLite database
dbWriteTable(conn, "merged_table", merged_df, overwrite = TRUE)
```

Print out tables and graphs for each year to see whether older planes suffers more delays.
```{r}
# Create an empty list to store ggplot objects
plot_list <- list()

for (year in 1990:1999) {
  age_query <- paste("
    SELECT 
      '", year, "' AS Year,
      manufactured_year AS manuyear,
      '", year, "' - manufactured_year AS age_of_plane,
      AVG((DepDelay + ArrDelay) / 2) AS avg_delay
    FROM merged_Table
    WHERE ArrDelay IS NOT NULL AND DepDelay IS NOT NULL
    AND manufactured_year <= ", year, " 
    GROUP BY age_of_plane
    ORDER BY age_of_plane ASC
  ")
  
  # Execute the query
  age_result <- dbGetQuery(conn, age_query)
  
  # print each year tables
  print(age_result)
  
  # Ensure the variable is numeric
  age_result$age_of_plane <- as.numeric(age_result$age_of_plane)
  
  # Create a ggplot object
  p <- ggplot(age_result, aes(x = age_of_plane, y = avg_delay)) +
    geom_line() + 
    labs(title = paste("Average Delay by Plane Age for Year", year),
         x = "Age of Plane", y = "Average Delay")
  
  # Store the ggplot object in the list
  plot_list[[as.character(year)]] <- p
}

# Print the line plots for each year
for (year in 1990:1999) {
  print(plot_list[[as.character(year)]])
}
```
Disconnect from the database.
```{r}
dbDisconnect(conn)
```

part c
Load relevant packages
```{r}
library(dplyr)
```

Create dataframes to store probabilities and coefficients
```{r}
prob_df <- data.frame(Year = integer(), Probability = numeric())
coeff_df <- data.frame(Year = integer(), Feature = character(), Coefficient = numeric())
```

Create a dataframe for carriers.csv
```{r}
carriers_df <- read.csv("carriers.csv", header = TRUE)
```

Create tables and barplots for each year then fit logistic regression model
```{r}
# Loop over each year
for (i in c(1990:1999)) {
  # Filter data for each year
  ontime_c <- read.csv(paste0(i, ".csv"), header = TRUE)
  
  # Left join with carriers_df to include all unique carrier information
  ontime_cc <- ontime_c %>%
    left_join(carriers_df, by = c("UniqueCarrier" = "Code"))
  
  # Prepare the features and target variable which is "Diverted"
  features <- c("DayOfWeek", "CRSDepTime", "DepTime","CRSArrTime",  "DepDelay", "Distance", "UniqueCarrier")
  target <- "Diverted"
  
  # Convert "UniqueCarrier" from character to a factor
  ontime_cc$UniqueCarrier <- as.factor(ontime_cc$UniqueCarrier)
  
  # Fit logistic regression model
  model <- glm(formula = as.formula(paste(target, "~", paste(features, collapse = " + "))),
                data = ontime_cc, family = binomial)
 
  # Predict probabilities of plane being diverted
  probabilities <- predict(model, type = "response")
  
  # Aggregate and store probabilities
  year_prob <- mean(probabilities)
  prob_df <- rbind(prob_df, data.frame(Year = i, Probability = year_prob))

  # Get coefficients
  coefficients <- coef(model)
  
  # Store coefficients and its corresponding year
  coeff_df <- rbind(coeff_df, data.frame(Year = rep(i, length(coefficients)), Feature = names(coefficients), Coefficient = coefficients))

  # Sort coefficients in ascending order 
  coeff_sorted <- coefficients[order(coefficients)]
  
  # Create a barplot for each year
  barplot(coeff_sorted, main = paste("Logistic Regression Coefficients for Year", i),
          xlab = "Features", ylab = "Coefficient Value")
  
  # Print coeff_df for each year
  cat("Coefficients for year", i, "in ascending order:\n")
  print(data.frame(Feature = names(coeff_sorted), Coefficient = coeff_sorted))
}

# Plot a line graph for probability of diversion over 1990 to 1999
plot(prob_df$Year, prob_df$Probability, type = "l",
     main = "Probability of Diverted US Flights Over Time",
     xlab = "Year", ylab = "Probability")

```