---
title: "part 1"
author: "220455224"
date: "`r Sys.Date()`"
output: html_document
---
QUESTION 1A)
set up initial values 
x0 is the initial value
N is the number of iterations
s is the standard deviation
```{r}
x0 <- 0 
N <- 10000
s <- 1
```
Set up Metropolis-Hastings algorithm
```{r}
metropolis_hastings <- function(initial_value, iterations, std_dev){
  samples <- numeric(iterations)
  current_value <- initial_value
  
  for (i in 1:iterations) {
    x_asterisk <- rnorm(1, mean = current_value, sd = std_dev)
    
    log_ratio <- -abs(x_asterisk) + abs(current_value)
    log_u <- log(runif(1))
    
    if (log_u < log_ratio) {
      current_value <- x_asterisk
    }
    
    samples[i] <- current_value
  }
  
  return(samples)
}
```
Run the metropolis hasting algorithm
```{r}
samples <- metropolis_hastings(x0 , N , s)
```
Plot histogram and kernel plot density
```{r}
hist(samples, main= "Metropolis-Hastings Samples", xlab = "x", prob = TRUE)
lines(density(samples), col = "blue")
curve(0.5 * exp(-abs(x)), from = -10, to = 10, col = "red", add = TRUE)
```
Get sample mean and sample standard deviation
```{r}
cat("Sample Mean:" , mean(samples), "\n")
cat("Sample Standard Deviation:", sd(samples), "\n")
```
QUESTION 1B)
Set up Metropolis-Hastings algorithm
```{r}
metropolis_hastings <- function(initial_value, iterations, std_dev, j) {
  samples <- numeric(iterations)
  current_value <- initial_value

  for (i in 1:iterations) {
    x_star <- rnorm(1, mean = current_value, sd = std_dev)
    ratio <- exp(-(abs(x_star)^j)) / exp(-(abs(current_value)^j))
    u <- runif(1)

    if (u < ratio) {
      current_value <- x_star
    }

    samples[i] <- current_value
  }

  return(samples)
}
```

Set the parameters as given by the question
```{r}
N <- 2000
s <- 0.001
J <- 4
initial_x <- 0
```

Run the algorithm and print Mj and Vj values
```{r}
for (j in 1:J) {
  # Generate J chains with different initial values
  chains <- replicate(J, metropolis_hastings(initial_x, N, s, j), simplify = FALSE)

  # Calculate Mj for each chain
  Mj_values <- sapply(chains, function(chain) mean(chain^j))
  
  # Print mean of each chain
  cat("For j =", j, "\n")
  cat("Mj values:", Mj_values[j], "\n")

  # Calculate variance
  Vj <- sapply(chains, function(chain) var(chain^j))

  # Print Variance of each chain
  cat("Variance values:", Vj[j], "\n")
}
```
Calculate and print the overall within-sample mean W
```{r}
# Calculate the overall within-sample mean W 
W <- mean(Vj)

# Print W
cat("Overall Within-Sample Variance W:", W)
```
Calculate and print overall sample mean M
```{r}
# Calculate overall mean M
M <- mean(Mj_values)

# Print M 
cat("Overall Sample Mean M:", M)
```
Calculate and print between sample variance B
```{r}
# Calculate Between sample variance
B <- mean((Mj_values - M)^2)

# Print B
cat("Between Sample Variance B:", B)
```
Calculate and print R hat
```{r}
# Calculate R_hat
R_hat <- sqrt((B + W) / W)

# Print R_hat 
cat("R_hat:", R_hat)
```
When s-values is in the interval between 0.001 and 1

Set seed
```{r}
set.seed(1)
```
Calculate R hat for a given s
```{r}
calc_R_hat <- function(s) {
  # Set parameters
  N <- 2000
  J <- 4
  initial_x <- 0
  
  # Initialize vectors to store Mj and Vj values
  Mj_values <- numeric()
  Vj_values <- numeric()
  
  # Calculate Mj and Vj for each chain and collect them
  for (j in 1:J) {
    chains <- replicate(J, metropolis_hastings(initial_x, N, s, j), simplify = FALSE)
    
    # Calculate Mj for each chain
    Mj_values <- c(Mj_values, sapply(chains, function(chain) mean(chain^j)))
    
    # Calculate Vj for each chain
    Vj_values <- c(Vj_values, sapply(chains, function(chain) var(chain^j)))
    
    # Print mean of each chain
    cat("For j =", j, "\n")
    cat("Mj values:", Mj_values[j], "\n")
    # Print Variance of each chain
    cat("Variance values:", Vj_values[j], "\n")
  }
  
  # Calculate the overall within-sample mean W 
  W <- mean(Vj_values)
  
  # Calculate overall mean M
  M <- mean(Mj_values)
  
  # Calculate Between sample variance
  B <- mean((Mj_values - M)^2)
  
  # Calculate R_hat
  R_hat <- sqrt((B + W) / W)
  
  # Print results
  cat("Overall Within-Sample Variance:", W, "\n")
  cat("Overall Sample Mean:", M, "\n")
  cat("Between Sample Variance B:", B, "\n")
  cat("R_hat:", R_hat, "\n")
  
  return(R_hat)
}
```
Set s values by the given parameter of 0.001 to 1
```{r}
s_values <- seq(0.001, 1, length.out = 100)
```
Calculate and print R hat for each s-value while also printing the chains values for checking
```{r}
mean_R_hat_values <- sapply(s_values, calc_R_hat)
```
Plot the graph
```{r}
plot(s_values, mean_R_hat_values, type = "l", xlab = "s", ylab = "R_hat", main = "R_hat over s values")
```